A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-03-01 analysis/paper/paper.Rmd
Slight adjustment to figures. Added a couple references

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; see also discussion in @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm, led the site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2012; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2011a]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227]. @nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also e.g. @amundsen2006; @solheim2012, 74; @ostmo1988]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.

@@

Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @åstveit2018; @solheim2020, see also @akerlund1995]. This approach has a couple of limitations. First of all, the displacement curves are commonly applied directly to larger study areas, with only a few studies having taken the variable uplift-rates into account when performing this comparison [e.g. @fossum2020; @møller1987; @persson2008]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating. Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.

@@

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, as indicated by the clear deviation in site location from the shoreline after this, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, while 95% of the values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m. That is, for 95% of the cases, the shoreline was situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location. This is constrained slightly when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m. While the median distances for horizontal and topographic is only and respectively, it also evident that the range of these values is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the topography.

An exponential function has been fit to the distributions for vertical distance using maximum likelihood. While it makes theoretical sense that a process of exponential decay explains this relationship, it  is clear that this does not perfectly match the data. However, this can in part be related to methodological factors, where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero. Having derived an exponential decay function for describing the vertical distance between sites and shoreline, this can be combined with the displacement data to provide a method for shoreline dating that takes this into account:

Where x is, asdf. This formula is used to shoreline date the same sites from where this relationship was derived in Figure \@ref(fig:shoredate). The Løvås sites and Gunnarsrød 5 deviate expectedly from the shoreline date. Nordby 52 is not equally expected to deviate from the shoreline date, but this could follow from the fact that a highway runs by the site and had to be edited. Using the method for evaluating degree of contemporaneity from @parnell2008, given in figure , indicates that the shoreline date is conteporenous date with the radiocarbon dates phase for.

Using the resulting exponential decay function, the sites treated here have been given a shoreline date

Following the general and aggregative result of this, one recommendation could be that given no other information, the method of shoreline dating should be applied with an uncertainty of 0-15m offset to the Stone Age site elevation relative to the displacement curves, following the median vertical distance and associated interquartile range. This would naturally be best applied when combined with a more comprehensive consideration of the site location, which could lead one to reduce this uncertainty, given for example surroundings where it would be reasonable to assume that a certain sea-levels would be unlikely in light of the surroundings of the site.

Figure shows the result of shoreline dating the sites using the recommended off-set with radiocarbon dates plotted against this. The steeper the curve the more narrow the date range resulting from the procedure. This means that older siter, and sites situated towards the north-east will have a more narrow date range.

@@

It is worth mentioning some sources of likely variation and uncertainty that have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, has not been taken into account. Secondly, the DTM has a vertical error [cf. @lewis2021].

Thirdly, the tidal range has not been considered...   And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site. These dimensions will have implications for the shore-line dating for certain subsets of the sites and could be future avenues through which one could further improve the method. These dimensions will also be imprtant for more substantial considerations of the behavioural relation between sites and the shoreline.

@@

Finally, the analysis done here was Monte Carlo-simulation, integrating multiple sources of spatio-temporal uncertainty. Here the approach was simply used to inform the question of the distance between sites and shorelines. However, this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

@@

 Furthermore, the typical application of shoreline dating as an approximate method where date ranges are often for example rounded off to 200-year intervals, means that most previous research based on the method is given support by the findings reported here. However, the findings also suggest that simply using the method as a terminus ante quem for the occupation of the sites is, at least in my view, overly conservative and would stand to unnecessarily loose important chronological information. Attempting to move these considerations further through an explicit quantification has here resulted in a concrete recommendation of applying an offset of 0-15m when shoreline dating sites where no other information is available.

>>>>

The shore-bound settlement location of prehistoric hunter-fisher-gatherers in Norway is generally believed to follow both from the exploitation of aquatic resources and from movement and communication, which would have been efficient on waterways. The same logic has also been extended to the hinter- and inland regions, where sites are to be located along rivers and lakes [@brogger1905; @glorstad2010, 57--87; but see discussion in @gundersen2013; @mjærum2018; @schulke2020]. This is to take a dramatic turn at the transition to the Late Neolithic, around 2400 BCE, with the introduction of the Neolithic proper [@solheim2021; @prescott2020]. The introduction of a comprehensive Neolithic cultural package, including a shift to agro-pastoralism and the introduction of the farm, led the site locations to be more withdrawn from the shoreline [e.g. @bakka1971; @prescott2012; @ostmo2008, 223]. That is not to say that waterways and aquatic resources were no longer exploited, but rather that these activities would not have been as tightly integrated with settlement and tool-production areas as in preceding periods [@glorstad2011a]. At an earlier stage, at the transition to the Early Neolithic (c. 3900 BCE), pottery is introduced to the sites, and there are some indications of an initial uptake of agriculture at some sites in the Oslo fjord region. However, this appears to be small in scale and is believed to be combined with a continued and predominantly hunter-gatherer life-way, possibly followed by a complete de-Neolithisation in the Middle Neolithic [@nielsen2019; @hinsch1955; @ostmo1988, 225--227]. @nielsen2021b has recently argued that the initial uptake of agriculture in the Early Neolithic is combined with a more complex settlement pattern, and that a simple forager/farming dichotomy would underplay the variation present in the Early and Middle Neolithic data [see also e.g. @amundsen2006; @solheim2012, 74; @ostmo1988]. Seen in relation to the question of interest here, the empirical expectation for the above outlined development would thus be a predominantly shore-bound settlement in the Mesolithic, possibly followed by a more varied association between sites and the shore-line from around 3900 BCE, and a decisive shift around 2400 BCE.

@@

Other previous evaluations of the correspondence between radiocarbon- and RSL-informed dates have typically followed the same structure as that of @breivik2018, involving a visual inspection of radiocarbon probability density functions plotted against local shoreline displacement curves based on the elevation of the site [e.g. @åstveit2018; @solheim2020; see also @akerlund1995]. This approach has a couple of limitations. First of all, the displacement curves are commonly applied directly to larger study areas, with only a few studies having taken the variable uplift-rates into account when performing this comparison [e.g. @fossum2020; @møller1987; @persson2008]. Secondly, with this method, the wider the uncertainty range associated with either radiocarbon date or displacement curve, the higher the probability that the confidence intervals overlap, and the higher the probability that we conclude in favour of our hypothesis. This thus leads to an inferential framework that favours uncertainty, which is hardly desirable. In statistical terms this follows from the fact that while one cannot conclude that two dates are different if their confidence intervals overlap, this does not necessarily mean that they are the same. The question thus necessitates a flip from a null-hypothesis of no significant difference, to one of equivalence [e.g. @lakens2018], as the question of interest is effectively one of synchroneity between events [cf. @parnell2008]. Another limitation of this often-employed method is that it only takes into account the vertical distance between the sites and the sea-level. While this is the main parameter of interest for shoreline dating, the practical implications of a vertical difference in RSL will be highly dependent on local topography and bathymetry. RSL-change can have more dramatic consequences in a landscape characterised by a low relief, as the horizontal displacement of the shoreline will be greater. Taking the spatial nature of the relationship between site and shoreline into account will consequently help get more directly at the behavioural dimension of this relation, and move the analysis beyond a purely instrumental consideration of the applicability of shoreline dating. Suggested ways to help mitigate and integrate the issues presented above into to the analysis are presented in the methods section.

@@

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, as indicated by the clear deviation in site location from the shoreline after this, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. The resulting best point estimate for the vertical distance between site and shoreline for the pre-Late Neolithic is given by the median at `r round(median(distances2$vertdist))`m, while 95% of the values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m. That is, for 95% of the cases, the shoreline was situated on or down to `r round(quantile(distances2$vertdist, 0.95))`m below the site location. This is constrained slightly when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m. While the median for horizontal and topographic distances is only `r round(median(distances2$hordist))`m and `r round(median(distances2$vertdist))`m respectively, it is also evident that their range is substantially wider than that for vertical distance, illustrating the point that minor variations in vertical distance can have substantial consequences for these distance measures, depending on the topography.

An exponential function has been fit to the distributions for vertical distance using maximum likelihood. While it makes theoretical sense that a process of exponential decay explains this relationship, it is also clear that this does not perfectly match the data. However, this can at least in part be related to methodological factors where the accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero. If one accepts this, having derived an exponential decay function for describing the vertical distance between sites and shoreline can be combined with the displacement data to provide a method for shoreline dating that takes this distance into account:

Where x is [...] This formula is used to shoreline date the same sites from where this relationship was derived (Figure \@ref(fig:shoredate). The result of comparing the synchroneity of the dates using the method of @parnell2008, is given in figure \@ref(fig:shoredate)B. Here, 100,000 age samples drawn from each ^14^C-date was subtracted from 100 000 age samples from the corresponding shoreline date. The result of this indicates that the shoreline date correspond with the radiocarbon dates for 44 out of 49 sites `r round(44/49, 2)*100`%. The deviation from the shoreline date at the Løvås sites and Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langangen Vestgård 1 and Nordby 52. For these two sites the zero age difference can be seen to fall just outside the 95% interval, and in the simulation of the distance to the shoreline the shoreline was simulated to be 0m from the site across virtually all of the 1000 simulation runs for both sites (see supplementary material).

@@

It is worth mentioning some sources of likely variation and uncertainty that have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, has not been taken into account. Secondly, the DTM has a vertical error [cf. @lewis2021]. Thirdly, while mean sea-level is used for dating the sites, a consideration of the tidal range

  And finally, the nature of the visits to the sites will presumably also have implications for how exposed a site. These dimensions will have implications for the shore-line dating for certain subsets of the sites and could be future avenues through which one could further improve the method. These dimensions will also be imprtant for more substantial considerations of the behavioural relation between sites and the shoreline.

@@

Finally, this analysis was based on Monte Carlo-simulation for integrating multiple sources of spatio-temporal uncertainty. Here the approach was simply used to inform the question of the distance between sites and shorelines. However, this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

@@

ges are often for example rounded off to 200-year intervals, means that most previous research based on the method is given support by the findings reported here. However, the findings also suggest that simply using the method as a terminus ante quem for the occupation of the sites is overly conservative and would stand to loose important chronological information unnecessarily. Attempting to move these considerations further through an explicit quantification has here resulted in a concrete recommendation of applying an offset when shoreline dating sites where no other information is available.
